Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge

### 2019-07-08 man/ecosystems/ms_ecosystems.Rmd
update

The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017].

Aquí meter algo de lo que comentó RZ. Tree level response y stand response.

Mira la biblio de regino y esto https://www.researchgate.net/publication/327321775_Drought_Sensitiveness_on_Forest_Growth_in_Peninsular_Spain_and_the_Balearic_Islands

@@

Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warmer site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.

@@

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts were also reported. Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@ MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not climatic rear-edge.

A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability previously and after the drought event. For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.gfirst months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites). This could explain the higher values of recovery found for 1995 (Figure 6). The high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality occurred in this area (*personal communications*).

@@

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b]. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive [@Bhuyan2017].

@@

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].

@@

# Conclussions 
Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests at Sierra Nevada mountain range. However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed. Surprisingly we observed high resilience values to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.

The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019].

The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.

Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.

@@

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. We also are very thankful to F.J. Bonet-García for his valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.

>>>>

The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017].

Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlithg a stronger reponse of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.

@@

Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warmer site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope. Sampling sites follow a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.

@@

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts were also reported. Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not climatic rear-edge.

A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability previously and after the drought event. For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.g.* first months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites). This could explain the higher values of recovery found for 1995 (Figure 6). The high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality occurred in this area (*personal communications*).

@@

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b]. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.  [@Bhuyan2017].

@@

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada (Table S4). As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].

@@

# Conclussions 
Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests at Sierra Nevada mountain range. However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed. Surprisingly we observed high resilience values to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019]. The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain. Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.

@@

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. F.J. Cano-Manuel and F.J. Navarro provided worthwhile information about oak management projects in Sierra Nevada. We also are very thankful to F.J. Bonet-García for his valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.
